---
title: "DiRT v2 — Dicistronic Transcript Detection (optimized)"
output: html_document
date: "`r Sys.Date()`"
params:
  gff_file:         "Genome_annotation.gff3"
  genome_fai:       "Reference_genome.fasta.fai"
  tRNA_bed:         "tRNA.bed"
  bam_dir:          "bam"
  bam_pattern:      "\\.sorted\\.bam$"
  sample_names:     !r c("ERR1", "ERR2", "ERR3")
  min_count:        1
  fdr_threshold:    0.05
  out_dir:          "results"
  bedtools_dir:     "bedtools_work"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  # ---- Cache configuration -----------------------------------------------
  # Caching is enabled per-chunk via `cache = TRUE`. We only cache the
  # expensive BAM-reading chunks (count-tRNA, count-genes, intergenic-info).
  # autodep = TRUE means a cached chunk is automatically invalidated when
  # any upstream chunk it depends on changes.
  cache.path = file.path(params$out_dir, "_cache/"),
  cache.lazy = FALSE,   # large objects (coverage lists) shouldn't be lazy-loaded
  autodep    = TRUE
)
dir.create(params$out_dir,      showWarnings = FALSE, recursive = TRUE)
dir.create(params$bedtools_dir, showWarnings = FALSE, recursive = TRUE)
```

```{r libraries}
suppressPackageStartupMessages({
  library(tidyverse)
  library(magrittr)
  library(data.table)
  library(GenomicAlignments)
  library(rtracklayer)
  library(GenomicFeatures)
  library(txdbmaker)
  library(openxlsx)
})
```

# Part 0 — Helper functions

```{r helpers}
# ---- File I/O helpers -------------------------------------------------------
out_path <- function(name) file.path(params$out_dir, name)
bed_path <- function(name) file.path(params$bedtools_dir, name)

write_tsv_plain <- function(df, file) {
  write.table(df, file = file, quote = FALSE, sep = "\t",
              row.names = FALSE, col.names = FALSE)
}

write_xlsx_table <- function(df, file) {
  openxlsx::write.xlsx(df, file, overwrite = TRUE)
}

# ---- Clean GFF gene-IDs (drops gene:/CDS:/transcript: prefixes and trailing
#      transcript suffixes like .1 or _t1)
clean_gene_id <- function(x) {
  x %>%
    str_remove("^(gene:|CDS:|transcript:)") %>%
    str_remove("(\\.\\d+|_t\\d+.*)$")
}

# ---- Coverage calculation (one Cal_Coverage that works for all cases) -------
Cal_Coverage <- function(gr, bam) {
  grl <- split(gr, f = gr$ids)
  lapply(grl, function(x) {
    # When the same id maps to >1 range (e.g. multiple genes share a nearest
    # intron) we average across ranges per base position-by-position by
    # returning a list of integer vectors.
    if (length(x) == 1) {
      chr <- as.character(seqnames(x))
      rng <- seq.int(start(x), end(x))
      coverage(bam, param = ScanBamParam(which = x))[[chr]][rng]
    } else {
      lapply(seq_along(x), function(i) {
        chr <- as.character(seqnames(x)[i])
        rng <- seq.int(start(x)[i], end(x)[i])
        coverage(bam, param = ScanBamParam(which = x[i]))[[chr]][rng]
      })
    }
  })
}

# ---- Per-sample AveCov (+ optional nZero) summarised into a wide tibble ----
summarise_coverage <- function(cov_list, sample_names, with_nZero = FALSE) {
  per_sample <- lapply(sample_names, function(s) {
    vals <- cov_list[[s]]
    df <- lapply(vals, function(i) {
      # Flatten Rle / list-of-Rle into a numeric vector
      v <- if (is.list(i)) unlist(lapply(i, as.numeric)) else as.numeric(i)
      if (with_nZero) {
        data.frame(nZero  = sum(v == 0),
                   AveCov = if (length(v)) mean(v) else NA_real_)
      } else {
        data.frame(AveCov = if (length(v)) mean(v) else NA_real_)
      }
    }) |> dplyr::bind_rows()
    if (with_nZero) {
      colnames(df) <- c(paste0(s, "_nZero"), paste0(s, "_AveCov"))
    } else {
      colnames(df) <- paste0(s, "_AveCov")
    }
    df
  }) |> dplyr::bind_cols()

  per_sample$ids <- names(cov_list[[1]])
  dplyr::select(per_sample, ids, dplyr::everything())
}

# ---- Build a GRanges with an "ids" mcol used by Cal_Coverage ---------------
# Renames any column whose name collides with a GRanges-reserved metadata
# name (e.g. an extra "strand" column passed when ignore.strand = TRUE) so
# the resulting GRanges object validates.
make_gr <- function(df, chr, start, end, ids, seqinfo, strand = NULL) {
  df2 <- as.data.frame(df, stringsAsFactors = FALSE)
  df2[[chr]]   <- as.character(df2[[chr]])
  df2[[start]] <- as.numeric(df2[[start]])
  df2[[end]]   <- as.numeric(df2[[end]])
  df2[[ids]]   <- as.character(df2[[ids]])

  reserved <- c("seqnames", "ranges", "strand", "seqlevels", "seqlengths",
                "isCircular", "start", "end", "width", "element")
  consumed <- c(chr, start, end, strand)            # these are pulled by makeGRanges
  clash    <- base::setdiff(base::intersect(colnames(df2), reserved), consumed)
  if (length(clash) > 0) {
    new_names <- paste0(clash, "_orig")
    colnames(df2)[match(clash, colnames(df2))] <- new_names
  }

  args <- list(
    df2,
    keep.extra.columns      = TRUE,
    ignore.strand           = is.null(strand),
    seqinfo                 = seqinfo,
    seqnames.field          = chr,
    start.field             = start,
    end.field               = end,
    starts.in.df.are.0based = FALSE
  )
  if (!is.null(strand)) args$strand.field <- strand
  do.call(makeGRangesFromDataFrame, args)
}

# ---- One paired-t.test driver used by all 3 intron classes -----------------
run_intron_intergenic_ttest <- function(aveCoverage_long) {
  trna_to_gene <- aveCoverage_long %>%
    filter(grepl("t[rR][nN][aA]", ids), !is.na(aveCoverage)) %>%
    distinct(ids, .keep_all = TRUE) %>%
    dplyr::select(ids, id) %>%
    mutate(id = as.character(id)) %>%
    split(f = .$id) %>%
    keep(~ nrow(.x) == 1) %>%
    bind_rows()

  if (nrow(trna_to_gene) == 0) {
    return(list(IntergenicVsintron_closest_to_intergenic                  = tibble(),
                intron_2nd_closest_to_intergenicVsintron_closest_to_intergenic = tibble()))
  }

  res <- lapply(trna_to_gene$ids, function(x) {
    combo <- filter(trna_to_gene, ids == x)$id
    cov <- filter(aveCoverage_long, id == combo) %>%
      reshape2::acast(Sample ~ type, value.var = "aveCoverage")

    out <- tibble(ids = character(), id = character(), comparison = character(),
                  df = double(), V = double(), p = double())

    if (all(c("intron_2nd_closest_to_intergenic", "intron_closest_to_intergenic")
            %in% colnames(cov)) &&
        nrow(cov) >= 2) {
      tt <- try(t.test(log(cov[, "intron_closest_to_intergenic"] + 1),
                       log(cov[, "intron_2nd_closest_to_intergenic"] + 1),
                       paired = TRUE), silent = TRUE)
      if (!inherits(tt, "try-error"))
        out <- bind_rows(out, tibble(
          ids = x, id = combo,
          comparison = "intron_2nd_closest_to_intergenicVsintron_closest_to_intergenic",
          df = tt$parameter, V = tt$statistic, p = tt$p.value))
    }

    if (all(c("intergenic", "intron_closest_to_intergenic")
            %in% colnames(cov)) &&
        nrow(cov) >= 2) {
      tt <- try(t.test(log(cov[, "intron_closest_to_intergenic"] + 1),
                       log(cov[, "intergenic"] + 1),
                       paired = TRUE), silent = TRUE)
      if (!inherits(tt, "try-error"))
        out <- bind_rows(out, tibble(
          ids = x, id = combo,
          comparison = "IntergenicVsintron_closest_to_intergenic",
          df = tt$parameter, V = tt$statistic, p = tt$p.value))
    }
    out
  }) %>%
    bind_rows()

  if (nrow(res) == 0) {
    return(list(IntergenicVsintron_closest_to_intergenic                  = tibble(),
                intron_2nd_closest_to_intergenicVsintron_closest_to_intergenic = tibble()))
  }

  res %>%
    mutate(adjP = p.adjust(p, "bonferroni"),
           FDR  = p.adjust(p, "fdr")) %>%
    arrange(p) %>%
    split(f = .$comparison)
}

# ---- Pass/fail filter on the two comparisons -------------------------------
# fdr_threshold: FDR cutoff used to call a comparison "significant"
#   • Intergenic vs intron_closest         must be significant AND V < 0
#     (i.e. intergenic coverage is greater than the closest intron's)
#   • intron_2nd_closest vs intron_closest must be NOT significant, OR
#     significant in the opposite direction (V > 0)
filter_significant <- function(ttest_results, fdr_threshold = 0.05) {
  ig <- ttest_results[["IntergenicVsintron_closest_to_intergenic"]]
  in_ <- ttest_results[["intron_2nd_closest_to_intergenicVsintron_closest_to_intergenic"]]
  if (is.null(ig) || is.null(in_) || nrow(ig) == 0) return(character(0))
  intergenic_sign <- ig %>% filter(FDR < fdr_threshold, V < 0)
  intron_notsign  <- if (nrow(in_) == 0) tibble(id = character()) else
    in_ %>% filter((FDR < fdr_threshold & V > 0) | FDR >= fdr_threshold)
  if (nrow(in_) == 0) intergenic_sign$id
  else base::intersect(intron_notsign$id, intergenic_sign$id)
}

# ---- Continuous-coverage assessment ----------------------------------------
assess_continuous_coverage <- function(nZero_intergenic, stats_parid, final_sign,
                                       coverage_intergenic, sample_names) {
  cov_list <- nZero_intergenic %>%
    merge(stats_parid, by = "ids") %>%
    filter(id %in% final_sign) %>%
    mutate(continu.cov = rowSums(dplyr::select(., contains("_nZero"))) == 0)

  if (nrow(cov_list) == 0) {
    return(cov_list %>% mutate(combined.continu.cov = logical()))
  }

  # Per-base matrix per intergenic id (rows = base positions, cols = samples).
  # Each sample's per-position coverage vector is a different length per
  # intergenic id, so we can't use vapply with a fixed FUN.VALUE — lapply +
  # cbind is the correct construction.
  all_intergenic <- names(coverage_intergenic[[sample_names[1]]])
  coverage_samples <- lapply(all_intergenic, function(ig) {
    per_sample <- lapply(sample_names, function(s) {
      v <- coverage_intergenic[[s]][[ig]]
      # When an id mapped to multiple ranges, Cal_Coverage returns a list of
      # Rles for that id; concatenate them position-wise into a single vector.
      if (is.list(v)) v <- unlist(lapply(v, as.integer)) else v <- as.integer(v)
      v
    })
    # Guard against samples that returned empty / different-length vectors
    lens <- vapply(per_sample, length, integer(1))
    if (any(lens == 0) || length(unique(lens)) != 1) return(NULL)
    do.call(cbind, per_sample)
  })
  names(coverage_samples) <- all_intergenic

  # Only keep ids that have a usable matrix in coverage_samples; anything
  # missing or NULL would otherwise blow up rowSums(m).
  kept <- base::intersect(cov_list$ids, names(coverage_samples))
  kept <- kept[!vapply(coverage_samples[kept], is.null, logical(1))]

  if (length(kept) == 0) {
    return(cov_list %>% mutate(combined.continu.cov = NA))
  }

  combined <- lapply(coverage_samples[kept], function(m) {
    if (!is.matrix(m)) m <- matrix(m, ncol = 1)
    if (nrow(m) == 0) return(NA_integer_)
    sum(rowSums(m) == 0)
  }) %>%
    unlist() %>% as.data.frame() %>% rownames_to_column() %>%
    set_colnames(c("ids", "Combined.nZero"))

  left_join(cov_list, combined, by = "ids") %>%
    mutate(combined.continu.cov = !is.na(Combined.nZero) & Combined.nZero == 0) %>%
    dplyr::select(-Combined.nZero)
}
```

# Part 1 — Parse GFF / build gene + tRNA tables

```{r parse-gff, cache = TRUE}
gff_gr <- rtracklayer::import(params$gff_file)
gff_df <- as.data.frame(gff_gr)

extract_feature <- function(gff_df, ftype, id_col = "ID") {
  gff_df %>%
    filter(type == ftype) %>%
    mutate(across(where(is.list), ~ sapply(., paste, collapse = ","))) %>%
    dplyr::select(any_of(c("seqnames", "start", "end", "strand", "type", id_col))) %>%
    set_colnames(c("Chr", "start", "end", "strand", "type", "gene.id")) %>%
    mutate(gene.id = clean_gene_id(gene.id))
}

gene.information     <- extract_feature(gff_df, "gene", "ID")
CDS.info.sorted      <- extract_feature(gff_df, "CDS",  "ID")
exon.info.canonical  <- extract_feature(gff_df, "exon", "Parent")

# Aggregate CDS per gene
CDS.start.end.for.gene <- CDS.info.sorted %>%
  mutate(start = as.numeric(start), end = as.numeric(end)) %>%
  group_by(gene.id) %>%
  summarise(gene.chr = dplyr::first(Chr),
            gene.start = min(start),
            gene.end   = max(end),
            gene.strand = dplyr::first(strand),
            gene.biotype = "protein_coding",
            .groups = "drop") %>%
  dplyr::select(gene.chr, gene.start, gene.end, gene.id, gene.strand, gene.biotype) %>%
  arrange(gene.chr, gene.start)

# Bed-style inputs for bedtools closest
write_tsv_plain(CDS.start.end.for.gene, bed_path("CDS_info.txt"))
write_tsv_plain(gene.information,       bed_path("Gene_info.txt"))
```

```{r parse-tRNA}
# tRNA.bed produced upstream by tRNAscan-SE + awk reformat
tRNA.raw <- read_delim(params$tRNA_bed, delim = "\t", col_names = FALSE,
                       show_col_types = FALSE)
colnames(tRNA.raw) <- c("chr", "start", "end", "amino_acid", "score", "anticodon")

tRNA.fixed <- tRNA.raw %>%
  mutate(number = row_number(),
         tRNA_id    = paste0("tRNA", number, "-", amino_acid, anticodon),
         bed_start  = pmin(start, end),
         bed_end    = pmax(start, end),
         bed_strand = ifelse(start < end, "+", "-"),
         bed_score  = 0) %>%
  dplyr::select(chr, bed_start, bed_end, tRNA_id, bed_score, bed_strand)

write_tsv_plain(tRNA.fixed, bed_path("tRNA_infor.txt"))

# Genome size for bedtools
seqdf <- read_delim(params$genome_fai, delim = "\t", col_names = FALSE,
                    show_col_types = FALSE) %>%
  dplyr::select(X1, X2) %>% set_names(c("seqnames", "seqlengths"))

write_tsv_plain(seqdf %>% mutate(seqnames = as.character(seqnames)),
                bed_path("Reference_genome.txt"))

Seqinfo <- with(seqdf, Seqinfo(seqnames    = as.character(seqnames),
                               seqlengths  = seqlengths,
                               isCircular  = rep(FALSE, length(seqnames)),
                               genome      = "genomeBuild")) |> sortSeqlevels()
```

# Part 1b — Bedtools closest (shell-out)

We invoke the BEDTools closest commands directly from R (the original protocol
ran them in a separate terminal). If `bedtools` is not on PATH this chunk fails
loudly — install it or set `eval=FALSE` here and run the commands manually.

```{r bedtools-closest}
bt_dir <- params$bedtools_dir
shell_sort <- function(infile, outfile, k = c("-k1,1", "-k2,2n")) {
  system2("sort", c(k, shQuote(infile)),
          stdout = outfile, stderr = "")
}
shell_sort(bed_path("CDS_info.txt"),         bed_path("CDS_info.sorted.txt"))
shell_sort(bed_path("tRNA_infor.txt"),       bed_path("tRNA_sorted.txt"))
shell_sort(bed_path("Reference_genome.txt"), bed_path("Reference_genome_sorted.txt"),
           k = "-k1,1")

system2("bedtools",
  c("closest", "-id",
    "-a", bed_path("tRNA_sorted.txt"),
    "-b", bed_path("CDS_info.sorted.txt"),
    "-g", bed_path("Reference_genome_sorted.txt"),
    "-D", "ref"),
  stdout = bed_path("closest_tRNA_gene_upS.bed"))

system2("bedtools",
  c("closest", "-iu",
    "-a", bed_path("tRNA_sorted.txt"),
    "-b", bed_path("CDS_info.sorted.txt"),
    "-g", bed_path("Reference_genome_sorted.txt"),
    "-D", "ref"),
  stdout = bed_path("closest_tRNA_gene_downS.bed"))
```

# Part 2 — tRNA-gene combination table

```{r tRNA-gene-combo}
closest_cols <- c("chr", "tRNA.start", "tRNA.end", "tRNA.id", "tRNA.score", "tRNA.strand",
                  "gene.chr", "gene.start", "gene.end", "gene.id", "gene.strand",
                  "gene.biotype", "distance")

closest_tRNA_CDS <- bind_rows(
    read.delim(bed_path("closest_tRNA_gene_upS.bed"),   header = FALSE),
    read.delim(bed_path("closest_tRNA_gene_downS.bed"), header = FALSE)) %>%
  unique() %>%
  set_colnames(closest_cols) %>%
  filter(gene.start != -1) %>%
  dplyr::select(-gene.chr, -tRNA.score) %>%
  mutate(tRNAFirst       = tRNA.start < gene.start,
         tRNA.upstream   = (gene.strand == "+" & distance > 0) |
                           (gene.strand == "-" & distance < 0),
         interval.length = ifelse(tRNAFirst,
                                  (gene.start - 1) - (tRNA.end   + 1) + 1,
                                  (tRNA.start - 1) - (gene.end   + 1) + 1))

tRNA.CDS.combo <- closest_tRNA_CDS %>%
  filter(distance != 0, interval.length != 0, gene.start != -1) %>%
  mutate(tRNA.id = as.character(tRNA.id),
         gene.id = as.character(gene.id),
         id = paste(pmin(tRNA.id, gene.id), pmax(tRNA.id, gene.id), sep = "_")) %>%
  group_by(gene.id) %>% dplyr::slice_min(interval.length, n = 1, with_ties = FALSE) %>%
  ungroup()

write_tsv_plain(tRNA.CDS.combo, bed_path("tRNA_CDS_Combination.txt"))
```

# Part 3 — Expression counts & filter commonly expressed tRNAs / genes

```{r bam-discovery}
abs_bam_dir <- normalizePath(params$bam_dir, mustWork = FALSE)
cat("Looking for BAMs in:", abs_bam_dir, "\n")
cat("Files in dir:\n")
print(list.files(params$bam_dir))

bamfiles  <- list.files(params$bam_dir, pattern = params$bam_pattern,
                        full.names = TRUE)
# Map BAM basenames → user-friendly sample names from params$sample_names.
# Convention: a sample name "ERR1" matches "ERR1.sorted.bam".
SampleName <- basename(bamfiles)
# Index sample names → BAM file (deterministic order matters everywhere downstream)
expected   <- paste0(params$sample_names, ".sorted.bam")
match_idx  <- match(expected, SampleName)
if (anyNA(match_idx)) {
  stop(
    "Could not match all params$sample_names to BAM files.\n",
    "  bam_dir (resolved):       ", abs_bam_dir, "\n",
    "  bam_pattern:              ", params$bam_pattern, "\n",
    "  Expected BAM filenames:   ", paste(expected, collapse = ", "), "\n",
    "  Files matching pattern:   ", if (length(SampleName)) paste(SampleName, collapse = ", ") else "(none)", "\n",
    "  ALL files in bam_dir:     ", paste(list.files(params$bam_dir), collapse = ", "), "\n",
    "Fix: pass an absolute path for bam_dir, e.g. ",
    "params = list(..., bam_dir = '/mnt/f/.../bam_merged')")
}
bamfiles   <- bamfiles[match_idx]
SampleName <- params$sample_names
names(bamfiles) <- SampleName
cat("Matched", length(SampleName), "BAMs:\n")
print(bamfiles)
```

```{r count-tRNA, cache = TRUE}
tRNA.info <- tRNA.fixed %>%
  set_colnames(c("tRNA.chr", "tRNA.start", "tRNA.end", "tRNA.id",
                 "tRNA.score", "tRNA.strand")) %>%
  mutate(across(c(tRNA.start, tRNA.end), as.numeric)) %>%
  arrange(tRNA.chr, tRNA.start)

GRange_tRNA <- makeGRangesFromDataFrame(tRNA.info,
                                        keep.extra.columns = TRUE,
                                        seqinfo = Seqinfo,
                                        seqnames.field = "tRNA.chr",
                                        start.field    = "tRNA.start",
                                        end.field      = "tRNA.end",
                                        strand.field   = "tRNA.strand")

tRNA.counts <- tRNA.info %>%
  dplyr::select(tRNA.chr, tRNA.start, tRNA.end, tRNA.id, tRNA.strand)
for (s in SampleName) {
  bam <- readGAlignments(bamfiles[s], index = paste0(bamfiles[s], ".bai"),
                         use.names = TRUE,
                         param = ScanBamParam(which = GRange_tRNA))
  tRNA.counts[[s]] <- countOverlaps(GRange_tRNA, bam, ignore.strand = TRUE)
}

write_xlsx_table(tRNA.counts, out_path("tRNA_counts.xlsx"))

count_cols <- SampleName
tRNA.expressed.AllSample <- rowSums(tRNA.counts[, count_cols] >= params$min_count) ==
  length(count_cols)
common.tRNA <- tRNA.counts[tRNA.expressed.AllSample, ]
tRNA.expressed.CDS.combo <- filter(tRNA.CDS.combo, tRNA.id %in% common.tRNA$tRNA.id)
```

```{r count-genes, cache = TRUE}
gene.counts.base <- tRNA.expressed.CDS.combo %>%
  dplyr::select(chr, gene.start, gene.end, gene.id, gene.strand) %>%
  unique()

GRange_gene <- makeGRangesFromDataFrame(gene.counts.base,
                                        keep.extra.columns = TRUE,
                                        ignore.strand = TRUE,
                                        seqinfo = Seqinfo,
                                        seqnames.field = "chr",
                                        start.field    = "gene.start",
                                        end.field      = "gene.end")

gene.counts <- gene.counts.base
for (s in SampleName) {
  bam <- readGAlignments(bamfiles[s], index = paste0(bamfiles[s], ".bai"),
                         use.names = TRUE,
                         param = ScanBamParam(which = GRange_gene))
  gene.counts[[s]] <- countOverlaps(GRange_gene, bam, minoverlap = 10L,
                                    ignore.strand = TRUE)
}
write_xlsx_table(gene.counts, out_path("gene_counts.xlsx"))

gene.expressed.AllSample <- rowSums(gene.counts[, SampleName] >= params$min_count) ==
  length(SampleName)
common.gene <- gene.counts[gene.expressed.AllSample, ]
tRNA.CDS.expressed.combo <- filter(tRNA.expressed.CDS.combo,
                                   gene.id %in% common.gene$gene.id)
```

# Part 4 — Intron information

```{r intron-info}
exon.tab <- exon.info.canonical %>%
  mutate(start = as.numeric(start), end = as.numeric(end))

intron.counts <- exon.tab %>%
  group_by(gene.id) %>%
  summarise(exon_count = dplyr::n(),
            gene.chr   = dplyr::first(Chr),
            gene.start = min(start),
            gene.end   = max(end),
            gene.strand = dplyr::first(strand),
            .groups = "drop") %>%
  mutate(intron_count = exon_count - 1)

# Buckets used by both the analysis & summary table
genes.no.intron     <- intron.counts %>% filter(intron_count == 0)
genes.one.intron    <- intron.counts %>% filter(intron_count == 1)
genes.multi.intron  <- intron.counts %>% filter(intron_count >= 2)

# Intron coordinates
intron.info <- exon.tab %>%
  dplyr::select(chr = Chr, start, end, gene.id, gene.strand = strand) %>%
  group_by(gene.id) %>%
  arrange(gene.id, start) %>%
  mutate(intron.start = end + 1,
         intron.end   = dplyr::lead(start) - 1,
         gene.strand  = dplyr::first(gene.strand),
         chr          = dplyr::first(chr)) %>%
  filter(!is.na(intron.end), intron.start <= intron.end) %>%
  arrange(gene.id,
          ifelse(gene.strand == "+", intron.start, -intron.start)) %>%
  mutate(intron.num = row_number(),
         intron.id  = paste0(gene.id, ".intron.", intron.num)) %>%
  ungroup() %>%
  dplyr::select(chr, intron.start, intron.end, intron.id, gene.id, gene.strand)
```

# Part 5 — Run the 3 intron-class analyses through one engine

```{r intergenic-info, cache = TRUE}
input.data <- tRNA.CDS.expressed.combo
id <- input.data$id

intergenic.info <- tibble(
  chr           = input.data$chr,
  start         = ifelse(input.data$tRNAFirst, input.data$tRNA.end + 1,  input.data$gene.end + 1),
  end           = ifelse(input.data$tRNAFirst, input.data$gene.start - 1, input.data$tRNA.start - 1),
  strand        = "NA",
  ids           = input.data$id,
  gene.id       = input.data$gene.id,
  type          = "intergenic",
  gene.strand   = input.data$gene.strand,
  sameStrand    = input.data$tRNA.strand == input.data$gene.strand,
  tRNAFirst     = input.data$tRNAFirst,
  tRNA.upstream = input.data$tRNA.upstream) %>%
  mutate(length = end - start + 1)

GRange_intergenic <- intergenic.info %>%
  mutate(ids = paste0("intergenic_", ids)) %>%
  make_gr("chr", "start", "end", "ids", Seqinfo)

# Compute intergenic coverage once (used by all 3 classes)
coverage.intergenic <- lapply(SampleName, function(s) Cal_Coverage(GRange_intergenic, bamfiles[s]))
names(coverage.intergenic) <- SampleName

AC.nZero.intergenic <- summarise_coverage(coverage.intergenic, SampleName, with_nZero = TRUE)
AC.intergenic       <- AC.nZero.intergenic %>% dplyr::select(ids, contains("AveCov"))
nZero.intergenic    <- AC.nZero.intergenic %>% dplyr::select(ids, contains("_nZero"))
# Re-label intergenic ids back to combo ids
AC.intergenic$ids   <- gsub("^intergenic_", "", AC.intergenic$ids)
nZero.intergenic$ids <- gsub("^intergenic_", "", nZero.intergenic$ids)

intergenic.data <- intergenic.info %>%
  dplyr::select(chr, start, end, ids, gene.id, strand = gene.strand,
                type, length, sameStrand, tRNAFirst, tRNA.upstream) %>%
  mutate(id = ids,
         strand = as.character(strand),
         sameStrand = as.character(sameStrand),
         tRNAFirst  = as.character(tRNAFirst),
         tRNA.upstream = as.character(tRNA.upstream))
```

```{r intergenic-rename-coverage}
# Strip the "intergenic_" prefix from per-sample coverage element names so
# they align with the cleaned `ids` used by AC.intergenic / nZero.intergenic.
# Kept in its own (uncached) chunk so the expensive `intergenic-info` cache
# above stays valid across re-renders.
coverage.intergenic <- lapply(coverage.intergenic, function(cov) {
  names(cov) <- gsub("^intergenic_", "", names(cov))
  cov
})
```

```{r class-runner}
# Single function that drives a single intron-class through coverage → t-test
# → continuous-coverage filter. `intron_blocks` is a tibble with columns
# (chr, start, end, ids, gene.id, type) where `type` is one of
# "intron_closest_to_intergenic" or "intron_2nd_closest_to_intergenic".
run_class <- function(intron_blocks, intergenic_data, sample_names, bamfiles, seqinfo,
                      coverage_intergenic, AC_intergenic, nZero_intergenic,
                      fdr_threshold = 0.05) {

  # 1. Build GRanges per intron type, compute coverage, summarise
  AC_list <- lapply(unique(intron_blocks$type), function(tt) {
    gr <- intron_blocks %>% filter(type == tt) %>%
      make_gr("chr", "start", "end", "ids", seqinfo)
    cov <- lapply(sample_names, function(s) Cal_Coverage(gr, bamfiles[s]))
    names(cov) <- sample_names
    summarise_coverage(cov, sample_names, with_nZero = FALSE)
  })
  names(AC_list) <- unique(intron_blocks$type)
  AC.intron      <- bind_rows(AC_list)
  AC.final       <- bind_rows(AC.intergenic %>% rename_with(~., everything()), AC.intron)
  AC.final$ids   <- as.character(AC.final$ids)

  # 2. Build the parid frame (intron + intergenic, joined to combo id)
  intron_pair <- intron_blocks %>%
    inner_join(intergenic_data %>% dplyr::select(combo_id = id, gene.id),
               by = "gene.id") %>%
    transmute(chr, start, end, ids,
              gene.id, strand = NA_character_, id = combo_id,
              type,   length = end - start + 1,
              sameStrand = NA_character_,
              tRNAFirst  = NA_character_,
              tRNA.upstream = NA_character_)
  final_stats_parid <- bind_rows(intron_pair, intergenic_data) %>%
    distinct() %>% arrange(id)

  # 3. Long-form coverage table for the t-test driver
  stats_test_keys <- final_stats_parid %>%
    dplyr::select(ids, chr, start, end, strand, id, type, length)
  ave_all <- AC.final %>%
    left_join(stats_test_keys, by = "ids") %>%
    arrange(id) %>% distinct()
  ave_long <- ave_all %>%
    reshape2::melt(id.vars = c("ids", "chr", "start", "end", "type", "id",
                               "strand", "length"),
                   value.name = "aveCoverage", variable.name = "Sample") %>%
    mutate(Sample = gsub("_AveCov$", "", Sample)) %>%
    as_tibble() %>%
    arrange(id, Sample, type)

  ttest_res  <- run_intron_intergenic_ttest(ave_long)
  final_sign <- filter_significant(ttest_res, fdr_threshold = fdr_threshold)

  cont_cov <- assess_continuous_coverage(nZero_intergenic, final_stats_parid,
                                         final_sign, coverage_intergenic, sample_names)
  list(ttest = ttest_res, sign = final_sign,
       continuous = cont_cov, stats_parid = final_stats_parid)
}
```

## 5a. Multi-intron candidates (2+ introns)

```{r class-multi}
# Direction-aware intron selection — MUST match the original DiRT v2:
#   tRNAFirst == TRUE  (tRNA upstream of gene, intergenic gap at 5' end)
#       -> "closest"     = 1st intron by start  (rank 1 ascending)
#       -> "2nd closest" = 2nd intron by start  (rank 2 ascending)
#   tRNAFirst == FALSE (tRNA downstream of gene, intergenic gap at 3' end)
#       -> "closest"     = LAST intron by start (rank 1 descending)
#       -> "2nd closest" = second-to-last       (rank 2 descending)
# Picking the wrong-end introns silently changes the t-test outcomes and
# both adds spurious hits and drops real ones (compared to the original
# pipeline) for any multi-intron gene with tRNA downstream of the CDS.
gene_direction <- intergenic.data %>%
  dplyr::select(gene.id, tRNAFirst) %>%
  distinct()

multi_intron_blocks <- intron.info %>%
  filter(gene.id %in% genes.multi.intron$gene.id) %>%
  inner_join(gene_direction, by = "gene.id") %>%
  group_by(gene.id) %>%
  arrange(intron.start, .by_group = TRUE) %>%
  mutate(
    rank_asc  = row_number(),
    rank_desc = dplyr::n() - row_number() + 1,
    # When tRNAFirst is TRUE pick the smallest start; when FALSE pick the
    # largest start.
    eff_rank  = ifelse(as.logical(tRNAFirst), rank_asc, rank_desc),
    type      = case_when(eff_rank == 1 ~ "intron_closest_to_intergenic",
                          eff_rank == 2 ~ "intron_2nd_closest_to_intergenic",
                          TRUE          ~ NA_character_)) %>%
  ungroup() %>%
  filter(!is.na(type)) %>%
  transmute(chr,
            start   = intron.start,
            end     = intron.end,
            ids     = intron.id,
            gene.id,
            type)

multi_res <- run_class(multi_intron_blocks, intergenic.data, SampleName, bamfiles,
                       Seqinfo, coverage.intergenic, AC.intergenic, nZero.intergenic,
                       fdr_threshold = params$fdr_threshold)
write_xlsx_table(multi_res$continuous, out_path("Continuous_Coverage_multi_intron.xlsx"))
```

## 5b. One-intron candidates (self intron + nearest neighbour intron)

```{r class-one, eval = TRUE}
# self introns — restricted to tRNA-paired expressed candidates (matches the
# original DiRT v2 pipeline). Without this restriction Cal_Coverage would run
# over every one-intron gene in the GFF (~thousands) instead of only the
# ~100 candidates whose tRNA + mRNA actually have RNA-seq coverage.
self_intron <- intron.info %>%
  filter(gene.id %in% genes.one.intron$gene.id,
         gene.id %in% intergenic.data$gene.id) %>%
  transmute(chr, start = intron.start, end = intron.end,
            ids = intron.id, gene.id,
            type = "intron_closest_to_intergenic")

# nearby introns via bedtools closest
one_bed <- bed_path("One_intron.bed")
write_tsv_plain(self_intron %>% mutate(start = start - 1) %>%
                  dplyr::select(chr, start, end, ids, gene.id), one_bed)
all_introns_bed <- bed_path("ALL_introns_info.txt")
write_tsv_plain(intron.info %>%
                  filter(!gene.id %in% self_intron$gene.id) %>%
                  mutate(intron.start = intron.start - 1) %>%
                  dplyr::select(chr, intron.start, intron.end, intron.id,
                                gene.id, gene.strand), all_introns_bed)
# bedtools closest requires inputs sorted by chrom, then start
one_bed_sorted         <- bed_path("One_intron.sorted.bed")
all_introns_bed_sorted <- bed_path("ALL_introns_info.sorted.txt")
shell_sort(one_bed,         one_bed_sorted)
shell_sort(all_introns_bed, all_introns_bed_sorted)
one_closest <- bed_path("One_Intron_closest_Intron.bed")
system2("bedtools", c("closest",
                      "-a", one_bed_sorted,
                      "-b", all_introns_bed_sorted),
        stdout = one_closest)

nearby_one <- read.delim(one_closest, header = FALSE) %>%
  set_colnames(c("chr", "self.start", "self.end", "self.id", "gene.id",
                 "n.chr", "n.start", "n.end", "n.id", "n.gene.id", "n.strand")) %>%
  filter(n.start != -1) %>%
  transmute(chr = n.chr, start = n.start, end = n.end,
            ids = n.id, gene.id,
            type = "intron_2nd_closest_to_intergenic")

one_intron_blocks <- bind_rows(self_intron, nearby_one)
one_res <- run_class(one_intron_blocks, intergenic.data, SampleName, bamfiles,
                     Seqinfo, coverage.intergenic, AC.intergenic, nZero.intergenic,
                     fdr_threshold = params$fdr_threshold)
write_xlsx_table(one_res$continuous, out_path("Continuous_Coverage_one_intron.xlsx"))
```

## 5c. No-intron candidates (two nearest neighbour introns)

```{r class-none, eval = TRUE}
no_bed <- bed_path("RNA_No_intron.bed")
# Restrict to tRNA-paired expressed no-intron candidates (matches the
# original DiRT v2 pipeline). Without this filter Cal_Coverage runs over
# every no-intron gene in the GFF instead of only the ~100 candidates.
no_intron_genes <- intron.counts %>%
  filter(intron_count == 0,
         gene.id %in% intergenic.data$gene.id) %>%
  inner_join(CDS.start.end.for.gene, by = "gene.id") %>%
  transmute(chr = gene.chr.x,
            start = gene.start.x - 1,
            end   = gene.end.x,
            ids   = gene.id,
            gene.id,
            gene.strand = gene.strand.x)
# fall back: when CDS isn't available use coordinates from intron.counts
if (nrow(no_intron_genes) == 0) {
  no_intron_genes <- intron.counts %>%
    filter(intron_count == 0,
           gene.id %in% intergenic.data$gene.id) %>%
    transmute(chr = gene.chr, start = gene.start - 1, end = gene.end,
              ids = gene.id, gene.id, gene.strand)
}
write_tsv_plain(no_intron_genes %>%
                  dplyr::select(chr, start, end, ids, gene.id, gene.strand), no_bed)

# bedtools closest requires inputs sorted by chrom, then start
no_bed_sorted <- bed_path("RNA_No_intron.sorted.bed")
shell_sort(no_bed, no_bed_sorted)
no_closest <- bed_path("RNA_No_Intron_closest2_Introns.bed")
system2("bedtools", c("closest",
                      "-a", no_bed_sorted,
                      "-b", all_introns_bed_sorted,
                      "-k", "2"),
        stdout = no_closest)

nearby_no <- read.delim(no_closest, header = FALSE) %>%
  set_colnames(c("chr", "self.start", "self.end", "self.id", "gene.id", "self.strand",
                 "n.chr", "n.start", "n.end", "n.id", "n.gene.id", "n.strand")) %>%
  filter(n.start != -1) %>%
  group_by(gene.id) %>%
  mutate(rank = row_number(),
         type = case_when(rank == 1 ~ "intron_closest_to_intergenic",
                          rank == 2 ~ "intron_2nd_closest_to_intergenic",
                          TRUE      ~ NA_character_)) %>%
  ungroup() %>%
  filter(!is.na(type)) %>%
  transmute(chr = n.chr, start = n.start, end = n.end,
            ids = n.id, gene.id, type)

no_res <- run_class(nearby_no, intergenic.data, SampleName, bamfiles,
                    Seqinfo, coverage.intergenic, AC.intergenic, nZero.intergenic,
                    fdr_threshold = params$fdr_threshold)
write_xlsx_table(no_res$continuous, out_path("Continuous_Coverage_no_intron.xlsx"))
```

# Part 6 — Combine all classes & classify localization

```{r combine-classify}
continuous.coverage.final.all <- bind_rows(
  multi_res$continuous %>% mutate(num.Intron = "2+"),
  one_res$continuous   %>% mutate(num.Intron = "1"),
  no_res$continuous    %>% mutate(num.Intron = "0")) %>%
  filter(combined.continu.cov == TRUE)

# Re-attach tRNA coordinates (using tRNA.fixed for full info)
tRNA_lookup <- tRNA.fixed %>%
  set_colnames(c("tRNA.chr", "tRNA.start", "tRNA.end", "tRNA.id",
                 "tRNA.score", "tRNA.strand"))

Dicistronic.all.tRNA <- continuous.coverage.final.all %>%
  left_join(tRNA.CDS.expressed.combo, by = "id") %>%
  dplyr::select(id,
                chr           = chr.x,
                start, end,
                gene.id       = gene.id.x,
                CDS.start     = gene.start,
                CDS.end       = gene.end,
                gene.strand,
                tRNA.id, tRNA.start, tRNA.end, tRNA.strand,
                tRNA.upstream = tRNA.upstream.x,
                tRNAFirst     = tRNAFirst.x,
                interval.length,
                continu.cov,
                num.Intron,
                combined.continu.cov,
                dplyr::contains("_nZero"))

DT_with_Gene_Startend <- Dicistronic.all.tRNA %>%
  left_join(gene.information %>%
              dplyr::select(gene.id, gene.start = start, gene.end = end),
            by = "gene.id")

DT_Final_Classified <- DT_with_Gene_Startend %>%
  mutate(tRNA_in_Gene = tRNA.start >= gene.start & tRNA.end <= gene.end,
         tRNA_in_CDS  = tRNA.start >= CDS.start & tRNA.end <= CDS.end,
         Location_Type = case_when(
           tRNA_in_CDS                                                  ~ "CDS-Internal",
           tRNA_in_Gene & !tRNA_in_CDS                                  ~ "UTR",
           (tRNA.start < gene.end & tRNA.end > gene.start) & !tRNA_in_Gene ~ "UTR",
           TRUE                                                         ~ "Intergenic"))

# Full result, still includes _nZero columns (preserves original behaviour)
write_xlsx_table(DT_Final_Classified, out_path("Final_Result_withIntronNmb.xlsx"))
```

# Part 7 — Final deliverables: Table 1 (summary) and Table 2 (clean DT table)

```{r table1-summary}
n_total       <- nrow(DT_Final_Classified)

# Intron breakdown from num.Intron
n_intronless  <- sum(DT_Final_Classified$num.Intron == "0", na.rm = TRUE)
n_single      <- sum(DT_Final_Classified$num.Intron == "1", na.rm = TRUE)
n_multi       <- sum(DT_Final_Classified$num.Intron == "2+", na.rm = TRUE)

# Localization buckets (UTR == UTR / Gene Boundary)
n_intergenic  <- sum(DT_Final_Classified$Location_Type == "Intergenic", na.rm = TRUE)
n_utr         <- sum(DT_Final_Classified$Location_Type == "UTR", na.rm = TRUE)
# (CDS-Internal would be reported separately if present.)

Table1_summary <- tibble(
  `Transcript Architectural Feature` = c(
    "Total Transcripts Detected",
    "Intron feature Breakdown",
    "— Intronless Genes",
    "— Single-Intron Genes",
    "— Multi-Intron Genes (2+)",
    "Localization Breakdown",
    "— Intergenic",
    "— UTR / Gene Boundary"),
  Count = c(
    n_total,
    NA_integer_,
    n_intronless,
    n_single,
    n_multi,
    NA_integer_,
    n_intergenic,
    n_utr))

write_xlsx_table(Table1_summary, out_path("Table1_Summary_DT_byFeature.xlsx"))
Table1_summary
```

```{r table2-clean}
# Keep the per-sample intergenic gap-base-pair counts, but rename them
# from `<sample>_nZero` to `<sample>_Intergenic_gap` for a more
# publication-friendly column header.
Table2_DT_clean <- DT_Final_Classified %>%
  dplyr::rename_with(~ sub("_nZero$", "_Intergenic_gap", .x),
                     .cols = dplyr::matches("_nZero$"))

write_xlsx_table(Table2_DT_clean, out_path("Table2_DT_Final_clean.xlsx"))
head(Table2_DT_clean)
```

# Session info

```{r session-info}
sessionInfo()
```
