---
title: "DiRT v2 — Dicistronic Transcript Detection (manually optimized, parameterized)"
output: html_document
date: "`r Sys.Date()`"
params:
  gff_file:        "annotation.gff3"
  genome_fai:      "genome.fasta.fai"
  tRNA_bed:        "tRNA.bed"
  bam_dir:         "bam"
  bam_pattern:     "\\.sorted\\.bam$"
  sample_names:    !r c("sample1","sample2","sample3")
  min_count:       1
  fdr_threshold:   0.05
  out_dir:         "results"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

####====================#### 
Packages Setup
####====================#### 

```{r}
library(tidyverse)
library(magrittr)
library(data.table)
library(GenomicAlignments)
library(rtracklayer)
library(GenomicFeatures)
library(txdbmaker)
library(reshape2)
library(biomaRt)
library(writexl)
library(GeneOverlap)
```

```{r setup-params, include=FALSE}
# Make YAML params visible to bash chunks via environment variables, and
# create the output directory. R chunks reference params$... directly.
Sys.setenv(GFF_FILE = params$gff_file,
           FAI_FILE = params$genome_fai,
           TRNA_BED = params$tRNA_bed,
           BAM_DIR  = params$bam_dir,
           OUT_DIR  = params$out_dir)
dir.create(params$out_dir, showWarnings = FALSE, recursive = TRUE)
```

####====================#### 
Part1: tRNA-gene combination from gene annotation files
####====================####

# Build canonical_transcripts (longest transcript per gene from the GFF).
# This uses the SAME GFF that is parsed below for gene.info.from.gff, so the
# canonical set is computed against the exact annotation used downstream.
# Same rule as the optimized version: slice_max(tx_len) per gene_id.
```{r build-canonical-transcripts}
Gff_tran_gr        <- rtracklayer::import(params$gff_file)
mcols(Gff_tran_gr) <- mcols(Gff_tran_gr)[, c("ID", "Parent", "type")]
Gff_tran_db        <- txdbmaker::makeTxDbFromGRanges(Gff_tran_gr)
tran_length_table  <- GenomicFeatures::transcriptLengths(Gff_tran_db)
canonical_transcripts <- tran_length_table %>%
  group_by(gene_id) %>%
  slice_max(order_by = tx_len, n = 1, with_ties = FALSE) %>%
  pull(tx_name)
```

### Gene information from GFF file
```{r}
gene.info.from.gff <- rtracklayer::import(params$gff_file) %>%
  as.data.frame()

gene.information <- gene.info.from.gff %>%
  filter(type == "gene") %>%
  mutate(across(where(is.list), ~ sapply(., paste, collapse = ","))) %>%
  dplyr::select(where(~ !all(is.na(.) | as.character(.) == "" | as.character(.) == "character(0)"))) %>%
  dplyr::select(c("seqnames", "start", "end", "strand", "type", "ID" )) %>%
  set_colnames(c("Chr", "start", "end", "strand", "type", "gene.id" ))%>%
  mutate(
    gene.id = str_remove(gene.id, "^(gene:|CDS:|transcript:)"),
    gene.id = str_remove(gene.id, "(\\.\\d+|_t\\d+.*)$")
  )

CDS.info.sorted <- gene.info.from.gff %>%
  filter(type == "CDS") %>%
  mutate(across(where(is.list), ~ sapply(., paste, collapse = ","))) %>%
  dplyr::select(where(~ !all(is.na(.) | as.character(.) == "" | as.character(.) == "character(0)")))%>%
  dplyr::select(c("seqnames", "start", "end", "strand", "type", "ID" )) %>%
  set_colnames(c("Chr", "start", "end", "strand", "type", "gene.id" )) %>%
  mutate(
    gene.id = str_remove(gene.id, "^(gene:|CDS:|transcript:)"),
    gene.id = str_remove(gene.id, "(\\.\\d+|_t\\d+.*)$")
  )
  
exon.info.canonical <- gene.info.from.gff %>%
  filter(type == "exon") %>%
  mutate(across(where(is.list), ~ sapply(., paste, collapse = ","))) %>%
  dplyr::select(where(~ !all(is.na(.) | as.character(.) == "" | as.character(.) == "character(0)")))%>%
  dplyr::select(c("seqnames", "start", "end", "strand", "type", "Parent" ))%>%
  set_colnames(c("Chr", "start", "end", "strand", "type", "gene.id" )) %>%
  # Strip optional 'gene:/CDS:/transcript:' prefix first so we can match against
  # canonical_transcripts (which holds transcript IDs like geneA_t1).
  mutate(gene.id = str_remove(gene.id, "^(gene:|CDS:|transcript:)")) %>%
  filter(gene.id %in% canonical_transcripts) %>%
  # Now collapse the _t<digit> suffix so we have one entry per gene.
  mutate(gene.id = str_remove(gene.id, "(\\.\\d+|_t\\d+.*)$"))

# Get CDS start and end points per gene
CDS.start.end.for.gene <- CDS.info.sorted %>%
  mutate(
    gene.start = as.numeric(start),
    gene.end = as.numeric(end)
  ) %>%
  group_by(gene.id) %>%
  dplyr::summarise(
    gene.id = dplyr::first(gene.id),
    gene.chr = dplyr::first(Chr),
    gene.start = min(gene.start),
    gene.end = max(gene.end),
    gene.strand = dplyr::first(strand),
    gene.biotype = dplyr::first(type),
    .groups = "drop"
  ) %>%
  dplyr::select(gene.chr, gene.start, gene.end, gene.id, gene.strand, gene.biotype) %>%
  arrange(gene.chr, gene.start)
```
# output the gene info for bedtools use
```{r}
write.table(CDS.start.end.for.gene,
            file = "CDS_info.txt",
            quote = FALSE, sep = "\t",
            row.names = FALSE, col.names = FALSE) 
```

# tRNA info from tRNAscan_SE
```{bash}
# Original block: cd + tRNAscan-SE + awk to produce tRNA.bed.
# `cd` removed (knit_root_dir is set by the Rscript caller). tRNAscan-SE
# skipped because the equivalent VvT2T_tRNA.bed is already provided in
# Input_Files and staged into the working directory as tRNA.bed by the
# caller. So this block is a no-op confirmation.
ls -la "$TRNA_BED"
```
# Read tRNA info
```{r}
tRNA.info <- read_delim(params$tRNA_bed, delim = "\t", col_names = FALSE)
colnames(tRNA.info) <- c("chr", "start", "end", "amino_acid", "score", "anticodon")
# Add a tRNA id
tRNA.info <- tRNA.info %>%
  dplyr::mutate(
    number = dplyr::row_number(),
    tRNA_id = paste0("tRNA", number, "-", amino_acid, anticodon))

#  Fix the coordinates and Create Strand Column
tRNA.fixed <- tRNA.info %>%
  mutate(
    # Create valid BED Coordinates (Always Min first, Max second)
    bed_start = pmin(start, end),
    bed_end   = pmax(start, end),
    
    # Create the Strand Column based on the original direction
    # If original Start > End, it is Negative (-)
    bed_strand = ifelse(start < end, "+", "-"),
    
    # Placeholder score (BED requires a score in col 5 usually, 0 is fine)
    bed_score = 0
  ) %>%
  #  Select exactly the 6 columns required for BED
  # Order: Chr, Start, End, Name, Score, Strand
  dplyr::select(chr, bed_start, bed_end, tRNA_id, bed_score, bed_strand)
```
# Output tRNA infor
```{r}
write.table(tRNA.fixed,
            file = "tRNA_infor.txt",
            quote = FALSE, sep = "\t",
            row.names = FALSE, col.names = FALSE)
```

# Make genome size file for bedtools use
```{bash}
# Original block: conda + samtools faidx + cut + sort + bedtools closest.
# `samtools faidx` skipped (T2T_ref.fasta.fai already provided in Input_Files
# and staged into the working directory). `conda activate` lines removed because
# they break a non-interactive bash chunk. Everything else is unchanged.
cut -f1,2 "$FAI_FILE" > Reference_genome.txt

# Sort the CDS file
sort -k1,1 -k2,2n CDS_info.txt > CDS_info.sorted.txt
# Sort the tRNA file
sort -k1,1 -k2,2n tRNA_infor.txt > tRNA_sorted.txt
sort -k1,1 Reference_genome.txt > Reference_genome_sorted.txt

bedtools closest -id -a tRNA_sorted.txt -b CDS_info.sorted.txt -g Reference_genome_sorted.txt -D ref > closest_tRNA_gene_upS.bed
bedtools closest -iu -a tRNA_sorted.txt -b CDS_info.sorted.txt -g Reference_genome_sorted.txt -D ref > closest_tRNA_gene_downS.bed
```
# combine two result files from BedTools and sort, then refine info into proper format
```{r}
closest_tRNA_CDS.up <- read.delim("closest_tRNA_gene_upS.bed", header=FALSE)
closest_tRNA_CDS.down <- read.delim("closest_tRNA_gene_downS.bed", header=FALSE)

closest_tRNA_CDS <- rbind(closest_tRNA_CDS.up,closest_tRNA_CDS.down) %>% 
  unique() %>% 
  set_colnames(c(
  "chr", "tRNA.start", "tRNA.end", "tRNA.id", "tRNA.score", "tRNA.strand",
  "gene.chr", "gene.start", "gene.end", "gene.id", "gene.strand", "gene.biotype",
  "distance")) 

closest_tRNA_CDS <- closest_tRNA_CDS %>%
  dplyr::filter(gene.start != -1) %>%
  dplyr::select(-gene.chr) %>%
  dplyr::select(-tRNA.score) %>%
  mutate(tRNAFirst = tRNA.start < gene.start,
         tRNA.upstream = ifelse(gene.strand == '+' & distance > 0, TRUE,
                                ifelse(gene.strand == '-' & distance < 0, TRUE, FALSE)),
         interval.length = ifelse(tRNAFirst == TRUE, 
                                  (gene.start-1)-(tRNA.end+1)+1, (tRNA.start-1)-(gene.end+1)+1)) 
```
## subset proper distance (not overlapped) for tRNA-gene combination and add unique id to them
# This pipeline is designed for non-stranded library, so only non-overlapped tRNA-gene combinations are considered as input. tRNA or gene id stand itself may cause issue for one match many, therefore, unique combination id is created.
```{r}
tRNA.CDS.combo <- closest_tRNA_CDS %>% 
  filter(distance != 0) %>%
  filter(interval.length != 0) %>%
  filter(gene.start != -1) %>%
  mutate(tRNA.id = as.character(tRNA.id),
         gene.id = as.character(gene.id)) %>% 
  split(f = seq(1, nrow(.))) %>% # break up the dataframe to one col one dataframe 
  lapply(function(x){mutate(x, id = paste(sort(c(x$tRNA.id, x$gene.id)), collapse = "_"))}) %>%
  bind_rows() %>% group_by(gene.id) %>% slice_min(n = 1, interval.length)%>%
    ungroup() 
```
# Output tRNA-gene combination info
```{r}
write.table(tRNA.CDS.combo,
            file = "tRNA_CDS_Combination.txt",
            quote = FALSE, sep = "\t",
            row.names = FALSE, col.names = FALSE)
```
# filter commonly expressed tRNA and gene 
# setup the bundary for the whole genome
```{r}
Seqinfo <- read_delim(params$genome_fai, 
                      delim = "\t", col_names = FALSE) %>%
  dplyr::select(X1, X2) %>%
  set_names(c("seqnames", "seqlengths"))

Seqinfo <- with(Seqinfo,
                Seqinfo(seqnames = seqnames,
                        seqlengths = seqlengths,
                        isCircular = rep(FALSE, length(seqnames)),
                        genome = "genomeBuild")) %>%
  sortSeqlevels()
```

####====================#### 
Part2: Count tRNA and gene expression
####====================####

## build up the ```GRange``` by ```makeGRangesFromDataFrame``` for tRNA
```{r}
GRange_tRNA <- makeGRangesFromDataFrame(tRNA.CDS.combo,
                                        keep.extra.columns = TRUE,
                                        ignore.strand = FALSE,
                                        seqinfo = Seqinfo,
                                        seqnames.field = "chr",
                                        start.field = "tRNA.start",
                                        end.field = "tRNA.end",
                                        strand.field = "tRNA.strand",
                                        starts.in.df.are.0based = FALSE)
```
## Count reads for each tRNA GRange from your RNA seq bam files 
```{r}
# specify Samples
bam.dir <- params$bam_dir
bamfiles <- list.files(bam.dir, pattern = params$bam_pattern, full.names = TRUE)
# Match BAM basenames to user-friendly sample names from params$sample_names
# (e.g. params$sample_names = c("RSL1","RSL2","RSL3","RSL4") and the BAM dir
# contains RSL1.sorted.bam, RSL2.sorted.bam, ...). The reordering below makes
# column order in count tables deterministic regardless of file-system order.
expected   <- paste0(params$sample_names, ".sorted.bam")
match_idx  <- match(expected, basename(bamfiles))
if (any(is.na(match_idx))) {
  stop("Could not match all params$sample_names to BAM files in ", bam.dir,
       "\n  Expected: ", paste(expected, collapse = ", "),
       "\n  Found:    ", paste(basename(bamfiles), collapse = ", "))
}
bamfiles    <- bamfiles[match_idx]
SampleName  <- params$sample_names
names(bamfiles) <- SampleName

counts <- lapply (bamfiles,function(bamfile){
  baifile <- paste0(bamfile,".bai")
  bam.input <- readGAlignments(bamfile,
                               index = baifile,
                               use.names = TRUE,
                               param = ScanBamParam(which = GRange_tRNA))
  
  countOverlaps(GRange_tRNA, bam.input,
                maxgap=0L, minoverlap=1L,
                type=c("any"),
                select=c("all"),
                ignore.strand=TRUE)
})

names(counts) <- SampleName
counts <- do.call(cbind,counts)
tRNA.counts <- cbind(tRNA.CDS.combo,counts) %>%
  dplyr::select(c("chr", "tRNA.start", "tRNA.end", "tRNA.id", "tRNA.strand", all_of(SampleName)))
```
## filter commonly expressed tRNA 
```{r}
# filter expressed tRNA 
tRNA.expressed.AllSample <- rowSums(dplyr::select(tRNA.counts, all_of(SampleName)) >= params$min_count) == length(SampleName)
# return logical value for all read count column
sum(tRNA.expressed.AllSample) 
common.tRNA <- tRNA.counts[tRNA.expressed.AllSample,]
tRNA.expressed.CDS.combo <- filter(tRNA.CDS.combo, tRNA.id %in% common.tRNA$tRNA.id) 
```
## build up the ```GRange``` by ```makeGRangesFromDataFrame``` for gene
```{r}
# prepare gene data
gene.counts <- tRNA.expressed.CDS.combo %>%
  dplyr::select(chr, gene.start, gene.end, gene.id, gene.strand) %>%
  unique()

# build up GRange for the gene
GRange_gene <- makeGRangesFromDataFrame(gene.counts,
                                        keep.extra.columns = TRUE,
                                        ignore.strand = TRUE,
                                        seqinfo = Seqinfo,
                                        seqnames.field = "chr",
                                        start.field = "gene.start",
                                        end.field = "gene.end",
                                        strand.field = "gene.strand",
                                        starts.in.df.are.0based = FALSE)
```
## Count reads for each gene GRange from ```bam``` files.
```{r}
counts <- lapply (bamfiles,function(bamfile){
  baifile <- paste0(bamfile,".bai")
  bam.input <- readGAlignments(bamfile,
                               index = baifile,
                               use.names = TRUE,
                               param = ScanBamParam(which = GRange_gene))
  
  countOverlaps(GRange_gene, bam.input,
                maxgap=0L, minoverlap=10L,
                type=c("any"),
                select=c("all"),
                ignore.strand=TRUE)
})
names(counts) <- SampleName
counts <- do.call(cbind,counts)
gene.counts <- cbind(gene.counts,counts)
```
## filter commonly expressed gene
```{r}
# filter expressed gene 
gene.expressed.AllSample <- rowSums(dplyr::select(gene.counts, all_of(SampleName)) >= params$min_count) == length(SampleName)
sum(gene.expressed.AllSample) 
common.gene <- gene.counts[gene.expressed.AllSample,]
tRNA.CDS.expressed.combo <- filter(tRNA.expressed.CDS.combo, gene.id %in% common.gene$gene.id)
```

####====================#### 
Part3: Get intron information
####====================####
# prepare exon information
```{r}
# define input.data
input.data <- tRNA.CDS.expressed.combo
# Add exon number for each exon
exon.info <- exon.info.canonical %>%
  dplyr::mutate(
    gene.start = as.numeric(start),
    gene.end   = as.numeric(end)
  ) %>%
  dplyr::group_by(gene.id) %>%
  dplyr::arrange(
    gene.id, 
    ifelse(strand == "+", gene.start, -gene.start)
  ) %>%
  dplyr::mutate(
    exon_UTR_id = paste0("exon", dplyr::row_number())
  ) %>%
  dplyr::ungroup() %>%
  dplyr::select(
    chr = Chr,
    start = gene.start,
    end = gene.end,
    exon_UTR_id,
    gene.id,
    strand
  )

# Filter for expressed genes
tgene.expressed.exon.info <- exon.info %>%
  filter(gene.id %in% input.data$gene.id)
```
#get how many tRNA.gene.expressed.combo have no intron
```{r}
### Fei edited: filter genes with no intron (one exon)
expressed.gene.exon_2 <- tgene.expressed.exon.info  %>% filter(`exon_UTR_id` == "exon2")
expressed.gene.only.one.exon <- subset(tgene.expressed.exon.info , !(gene.id %in% expressed.gene.exon_2$gene.id))

### filter positive result with only 2 exon (one intron)
tgene.expressed.exon3 <- tgene.expressed.exon.info %>% filter(`exon_UTR_id` == "exon3")
tgene.expressed.with1and2exon <- subset(tgene.expressed.exon.info, !(gene.id %in% tgene.expressed.exon3$gene.id)) 
tgene.expressed.with2exon <-  subset(tgene.expressed.with1and2exon, !(gene.id %in% expressed.gene.only.one.exon$gene.id))  %>% filter(`exon_UTR_id` == "exon2")
```
# pull out intron coordinates based on the exon information 
```{r}
# 1. Start with ONLY the Exon data
# Do not include UTR rows here, no reason to include them. 
intron.info <- exon.info %>%
  dplyr::select(
    chr = chr,
    start = start,
    end = end,
    gene.id,
    strand
  ) %>%
  dplyr::mutate(
    start = as.numeric(start),
    end = as.numeric(end)
  ) %>%
  dplyr::group_by(gene.id) %>%
  dplyr::arrange(gene.id, start) %>% 
  dplyr::mutate(
    intron.start = end + 1,
    intron.end = dplyr::lead(start) - 1,
    gene.strand = dplyr::first(strand), 
    chr = dplyr::first(chr)
  ) %>%
  dplyr::filter(!is.na(intron.end)) %>%
  dplyr::filter(intron.start <= intron.end) %>%
  dplyr::arrange(
    gene.id, 
    ifelse(gene.strand == "+", intron.start, -intron.start)
  ) %>%
  dplyr::mutate(
    intron.num = dplyr::row_number(),
    intron.id = paste0(gene.id, ".intron.", intron.num)
  ) %>%
  dplyr::ungroup() %>%
  dplyr::select(chr, intron.start, intron.end, intron.id, gene.id, gene.strand)
```
# prepare intron coverage data
Here, we had expressed tRNA-gene combination as potential candidates. Now, we need to assess the expression of intron and intergeneic region for each candidates to test if tRNA and gene are co-transcribed.
## Create GRanges to assess the coverage for each sample
### get intergenic and intron1&2 coordinates
```{r}
# keep combination id safe
id <- input.data$id

# get correct intergenic coordinates
intergenic.info <- data_frame(chr = input.data$chr,
                              start = ifelse(input.data$tRNAFirst == TRUE,
                                             input.data$tRNA.end+1,
                                             input.data$gene.end+1),
                              end = ifelse(input.data$tRNAFirst == TRUE,
                                           input.data$gene.start-1,
                                           input.data$tRNA.start-1),
                              strand = rep("NA", nrow(input.data)),
                              ids = input.data$id,
                              gene.id = input.data$gene.id,
                              type = rep("intergenic", nrow(input.data)),
                              gene.strand = input.data$gene.strand,
                              length = end-start+1,
                              sameStrand = input.data$tRNA.strand == input.data$gene.strand,
                              tRNAFirst = input.data$tRNAFirst,
                              tRNA.upstream = input.data$tRNA.upstream)

# get correct intron
# need closest and 2nd closest intron to intergenic
tRNA.up.intergenic <- filter(intergenic.info, 
                             tRNAFirst == TRUE)

tRNA.down.intergenic <- filter(intergenic.info, 
                               tRNAFirst == FALSE)

# corrdinates for closest intron to intergenic for tRNA.up.intergeic 
list.gene.id <- unique(tRNA.up.intergenic$gene.id)
res <- sapply(list.gene.id,function(g){
  rowNb <- which(intron.info$gene.id == g)
  start <- intron.info$intron.start[rowNb]
  or <- order(start)
  rowNb[or==1]
}) 
res <- unlist(res)

intron.closest.for.stats.up <- intron.info[res,]
intron.closest.for.stats.up <- data_frame(chr = intron.closest.for.stats.up$chr,
                                          start = intron.closest.for.stats.up$intron.start,
                                          end = intron.closest.for.stats.up$intron.end,
                                          strand = intron.closest.for.stats.up$gene.strand,
                                          ids = intron.closest.for.stats.up$intron.id,
                                          gene.id = intron.closest.for.stats.up$gene.id,
                                          type = rep("intron_closest_to_intergenic", nrow(intron.closest.for.stats.up)),
                                          gene.strand = intron.closest.for.stats.up$gene.strand,
                                          length = end-start+1,
                                          sameStrand = rep("NA", nrow(intron.closest.for.stats.up)),
                                          tRNAFirst = rep("NA", nrow(intron.closest.for.stats.up)),
                                          tRNA.upstream = rep("NA", nrow(intron.closest.for.stats.up)))

# range for 2nd closest intron to intergenic 
# if a gene dose not have more than one intron then the intron/intron comprison will be missing here
res <- lapply(list.gene.id,function(g){
  rowNb <- which(intron.info$gene.id == g)
  start <- intron.info$intron.start[rowNb]
  or <- order(start)
  c(rowNb[or==2]) 
}) 
res <- unlist(res)

intron.2nd_closest.for.stats.up <- intron.info[res,]
intron.2nd_closest.for.stats.up <- data_frame(chr = intron.2nd_closest.for.stats.up$chr,
                                              start = intron.2nd_closest.for.stats.up$intron.start,
                                              end = intron.2nd_closest.for.stats.up$intron.end,
                                              strand = intron.2nd_closest.for.stats.up$gene.strand,
                                              ids = intron.2nd_closest.for.stats.up$intron.id,
                                              gene.id = intron.2nd_closest.for.stats.up$gene.id,
                                              type = rep("intron_2nd_closest_to_intergenic", nrow(intron.2nd_closest.for.stats.up)),
                                              gene.strand = intron.2nd_closest.for.stats.up$gene.strand,
                                              length = end-start+1,
                                              sameStrand = rep("NA", nrow(intron.2nd_closest.for.stats.up)),
                                              tRNAFirst = rep("NA", nrow(intron.2nd_closest.for.stats.up)),
                                              tRNA.upstream = rep("NA", nrow(intron.2nd_closest.for.stats.up)))

# range for closest intron to intergenic for tRNA.down.intergeic 
list.gene.id <- unique(tRNA.down.intergenic$gene.id)
res <- sapply(list.gene.id,function(g){
  rowNb <- which(intron.info$gene.id == g)
  start <- intron.info$intron.start[rowNb]
  or <- order(start)
  rowNb[or==max(or)]
}) 
res <- unlist(res)

intron.closest.for.stats.down <- intron.info[res,]
intron.closest.for.stats.down <- data_frame(chr = intron.closest.for.stats.down$chr,
                                            start = intron.closest.for.stats.down$intron.start,
                                            end = intron.closest.for.stats.down$intron.end,
                                            strand = intron.closest.for.stats.down$gene.strand,
                                            ids = intron.closest.for.stats.down$intron.id,
                                            gene.id = intron.closest.for.stats.down$gene.id,
                                            type = rep("intron_closest_to_intergenic", nrow(intron.closest.for.stats.down)),
                                            gene.strand = intron.closest.for.stats.down$gene.strand,
                                            length = end-start+1,
                                            sameStrand = rep("NA", nrow(intron.closest.for.stats.down)),
                                            tRNAFirst = rep("NA", nrow(intron.closest.for.stats.down)),
                                            tRNA.upstream = rep("NA", nrow(intron.closest.for.stats.down)))


# range for 2nd closest intron to intergenic
res <- lapply(list.gene.id,function(g){
  rowNb <- which(intron.info$gene.id == g)
  start <- intron.info$intron.start[rowNb]
  or <- order(start)
  rowNb[or==max(or)-1] 
}) 
res <- unlist(res)

intron.2nd_closest.for.stats.down <- intron.info[res,]
intron.2nd_closest.for.stats.down <- data_frame(chr = intron.2nd_closest.for.stats.down$chr,
                                                start = intron.2nd_closest.for.stats.down$intron.start,
                                                end = intron.2nd_closest.for.stats.down$intron.end,
                                                strand = intron.2nd_closest.for.stats.down$gene.strand,
                                                ids = intron.2nd_closest.for.stats.down$intron.id,
                                                gene.id = intron.2nd_closest.for.stats.down$gene.id,
                                                type = rep("intron_2nd_closest_to_intergenic", nrow(intron.2nd_closest.for.stats.down)),
                                                gene.strand = intron.2nd_closest.for.stats.down$gene.strand,
                                                length = end-start+1,
                                                sameStrand = rep("NA", nrow(intron.2nd_closest.for.stats.down)),
                                                tRNAFirst = rep("NA", nrow(intron.2nd_closest.for.stats.down)),
                                                tRNA.upstream = rep("NA", nrow(intron.2nd_closest.for.stats.down)))


# combine info together for closest and 2nd closest intron
intron.closest <- do.call("rbind", list(intron.closest.for.stats.up,
                                        intron.closest.for.stats.down)) %>% 
  dplyr::arrange(gene.id) %>% 
  unique()

intron.2nd.closest <- do.call("rbind", list(intron.2nd_closest.for.stats.up,
                                            intron.2nd_closest.for.stats.down)) %>% 
  dplyr::arrange(gene.id) %>% 
  unique()
```

####====================#### 
Part4: Prepar t-test for candidates with one intron

As we know introns are the only sure regions that do not have reads, so that is a validate negative control of background reads and pre-mRNA reads. I will do t-test using introns from nearby two introns to represent introns from potential DT without intron or with one intron.
####====================####

### Step1. for candidates with 1 intron, make a bed file for their coordinates
```{r}
tRNA_mRNA_expressed_One_Intron <- tgene.expressed.with2exon %>%
  dplyr::mutate(
    start = as.numeric(start), 
    end = as.numeric(end)
  ) %>%
  dplyr::mutate(start = start - 1) %>% # Bedtools use 0 start position rather than other packages 1
  dplyr::select(chr, start, end, gene.id, strand) %>%
  set_colnames(c("gene.chr", "gene.start", "gene.end", "gene.id", "gene.strand"))%>%
  arrange(gene.chr, gene.start) %>%
  unique() 
# Remove introns from expressed genes with only one intron
introns.bed <- intron.info %>%
  dplyr::filter(!gene.id %in% tRNA_mRNA_expressed_One_Intron$gene.id) %>%
  dplyr::mutate(intron.start = intron.start - 1) %>%
  dplyr::select(chr, intron.start, intron.end, intron.id, gene.id, gene.strand)%>%
  arrange(chr, intron.start)
```
# Save the Poteintial DT candidates with 1/0 intron into bed files
```{r}
write.table(tRNA_mRNA_expressed_One_Intron,
            file = "One_intron.txt",
            quote = FALSE, sep = "\t",
            row.names = FALSE, col.names = FALSE)

# All introns information are in intron.info
write.table(introns.bed,
            file = "ALL_introns_info.txt",
            quote = FALSE, sep = "\t",
            row.names = FALSE, col.names = FALSE)
```

### Step2. Use Bed tools to get nearby gene with two or more introns for candidates with 1/0 intron
```{bash}
# For candidates with one intron — bedtools call unchanged.
bedtools closest -a One_intron.txt -b ALL_introns_info.txt > One_Intron_closest_Intron.bed
```
# Step3. Construct Intron information for candidates with one intron
# Load Closest introns information
```{r}
# tRNA-mRNA expressed candidates with one intron:
closest_Intron_for_tRNA_mRNA_expressed_One_Intron <- read.delim("One_Intron_closest_Intron.bed", header=FALSE) %>%
  set_colnames(c("chr", "One.intron.start", "One.intron.end", "One.intron.id", "One.intron.strand",
                 "nearby.Intron.Chr", "nearby.Intron.start", "nearby.Intron.end", "nearby.Intron.id","nearby.gene.id", "nearby.gene.strand"))  
```
# Make information for 1 intron genes
```{r}
# !!! The intron id here is two closes introns from the candidate gene
# One intron
One_intron_Nearby_Intron <- closest_Intron_for_tRNA_mRNA_expressed_One_Intron %>%
  dplyr::select(chr, nearby.Intron.start, nearby.Intron.end, nearby.Intron.id, One.intron.id) %>%
  mutate(type = "intron_2nd_closest_to_intergenic") %>%
  set_colnames(c("chr", "intron.start", "intron.end", "ids", "gene.id", "type")) 
```

# ONE intron, combine the intron from the gene and nearby gene together. Using its own intron as intron 1, and the nearby intron as intron 2.
```{r}
# Get their own intron
One_intron_Self_Intron <- intron.info %>%
  filter(gene.id %in% tRNA_mRNA_expressed_One_Intron$gene.id) %>%
  mutate(type = "intron_closest_to_intergenic")
  
# Get nearby intron
One_intron_Nearby_Intron_Clean <- One_intron_Nearby_Intron %>%
  left_join(
    One_intron_Self_Intron %>% dplyr::select(gene.id, gene.strand),
    by = "gene.id"
  )
# Make two new dataframe 
One_intron_Self_Intron_for_stats <- data_frame(chr = One_intron_Self_Intron$chr,
                                            start = One_intron_Self_Intron$intron.start,
                                            end = One_intron_Self_Intron$intron.end,
                                            strand = One_intron_Self_Intron$gene.strand,
                                            ids = One_intron_Self_Intron$intron.id,
                                            gene.id = One_intron_Self_Intron$gene.id,
                                            type = One_intron_Self_Intron$type,
                                            gene.strand = One_intron_Self_Intron$gene.strand,
                                            length = end-start+1,
                                            sameStrand = rep("NA", nrow(One_intron_Self_Intron)),
                                            tRNAFirst = rep("NA", nrow(One_intron_Self_Intron)),
                                            tRNA.upstream = rep("NA", nrow(One_intron_Self_Intron)))


One_intron_Nearby_Intron_for_stats <- data_frame(chr = One_intron_Nearby_Intron_Clean$chr,
                                            start = One_intron_Nearby_Intron_Clean$intron.start,
                                            end = One_intron_Nearby_Intron_Clean$intron.end,
                                            strand = One_intron_Nearby_Intron_Clean$gene.strand,
                                            ids = One_intron_Nearby_Intron_Clean$ids,
                                            gene.id = One_intron_Nearby_Intron_Clean$gene.id,
                                            type = One_intron_Nearby_Intron_Clean$type,
                                            gene.strand = One_intron_Nearby_Intron_Clean$gene.strand,
                                            length = end-start+1,
                                            sameStrand = rep("NA", nrow(One_intron_Nearby_Intron_Clean)),
                                            tRNAFirst = rep("NA", nrow(One_intron_Nearby_Intron_Clean)),
                                            tRNA.upstream = rep("NA", nrow(One_intron_Nearby_Intron_Clean)))

# When two DT share same closest gene, extract only unique introns and assign them to two DT genes later
# We use 'distinct' on the intron ID ('ids') so we don't have duplicates
unique_introns <- One_intron_Nearby_Intron_for_stats %>%
  dplyr::select(chr, start, end, ids) %>% # Keep only location info
  distinct(ids, .keep_all = TRUE)         # Remove duplicate rows like the one you found
```

####====================####
Part8: t-test for candidates without intron 0 intron
####====================####
# Step1. for candidates with 0 intron, make a bed file for their coordinates
```{r}
# One exon means no intron, load expressed genes with only one exon
tRNA_mRNA_expressed_No_Intron <- expressed.gene.only.one.exon %>%
  dplyr::mutate(start = start - 1) %>% 
  dplyr::select(chr, start, end, gene.id, strand) %>%
  set_colnames(c("gene.chr", "gene.start", "gene.end", "gene.id", "gene.strand"))%>%
  arrange(gene.chr, gene.start) %>%
  unique()
```
# Output
```{r}
write.table(tRNA_mRNA_expressed_No_Intron,
            file = "RNA_No_intron.txt",
            quote = FALSE, sep = "\t",
            row.names = FALSE, col.names = FALSE)
```
### Step2. Use Bed tools to get nearby gene with two or more introns for candidates with no intron

```{bash}
# For candidates with no intron — bedtools call unchanged.
bedtools closest -a RNA_No_intron.txt \
-b ALL_introns_info.txt \
-k 2 > RNA_No_Intron_closest2_Introns.bed
```
# Step3. Construct Intron information for candidates without intron
# tRNA-mRNA expressed candidates with No intron
```{r}
closest_Introns_for_tRNA_mRNA_expressed_No_Intron <- read.delim("RNA_No_Intron_closest2_Introns.bed", header=FALSE) %>%
  set_colnames(c("chr", "No.intron.start", "No.intron.end", "No.intron.id", "No.intron.strand",
                 "nearby.Intron.Chr", "nearby.Intron.start", "nearby.Intron.end", "nearby.Intron.id","nearby.gene.id", "nearby.gene.strand")) 
```
# Give a poper introns type for statistic
```{r}
No_intron_Nearby_Intron <- closest_Introns_for_tRNA_mRNA_expressed_No_Intron %>%
  dplyr::select(chr, nearby.Intron.start, nearby.Intron.end, nearby.Intron.id, No.intron.id, No.intron.strand) %>%
  set_colnames(c("chr", "intron.start", "intron.end", "ids", "gene.id", "gene.strand")) %>%
  dplyr::group_by(gene.id) %>%
  # Explicitly calling mutate ensures the internal data-masking for row_number() works
  dplyr::mutate(rank = dplyr::row_number()) %>% 
  dplyr::mutate(type = dplyr::case_when(
            rank == 1 ~ "intron_closest_to_intergenic",
            rank == 2 ~ "intron_2nd_closest_to_intergenic",
            TRUE      ~ "other_intron" 
          )) %>%
  dplyr::ungroup()
```

```{r}
No_intron_Nearby_Introns_for_stats <- data_frame(chr = No_intron_Nearby_Intron$chr,
                                                 start = No_intron_Nearby_Intron$intron.start,
                                                 end = No_intron_Nearby_Intron$intron.end,
                                                 strand = No_intron_Nearby_Intron$gene.strand,
                                                 ids = No_intron_Nearby_Intron$ids,
                                                 gene.id = No_intron_Nearby_Intron$gene.id,
                                                 type = No_intron_Nearby_Intron$type,
                                                 gene.strand = No_intron_Nearby_Intron$gene.strand,
                                                 length = end-start+1,
                                                 sameStrand = rep("NA", nrow(No_intron_Nearby_Intron)),
                                                 tRNAFirst = rep("NA", nrow(No_intron_Nearby_Intron)),
                                                 tRNA.upstream = rep("NA", nrow(No_intron_Nearby_Intron)))


# Separate the df into two for Grange count
No_intron_Nearby_Introns_for_stats_1st_nearby <- No_intron_Nearby_Introns_for_stats %>%
  filter(type == "intron_closest_to_intergenic")
No_intron_Nearby_Introns_for_stats_2st_nearby <- No_intron_Nearby_Introns_for_stats %>%
  filter(type == "intron_2nd_closest_to_intergenic")
```

# THE GRAND MERGE: Unifying all candidates for a single
```{r}
# 1. Combine all "1st Introns" (Type = intron_closest_to_intergenic)
Master_Intron_1st <- do.call("rbind", list(
  intron.closest,                                 # From >= 2 introns
  One_intron_Self_Intron_for_stats,               # From 1 intron
  No_intron_Nearby_Introns_for_stats_1st_nearby   # From 0 introns
)) %>% unique()

# 2. Combine all "2nd Introns" (Type = intron_2nd_closest_to_intergenic)
Master_Intron_2nd <- do.call("rbind", list(
  intron.2nd.closest,                             # From >= 2 introns
  One_intron_Nearby_Intron_for_stats,             # From 1 intron
  No_intron_Nearby_Introns_for_stats_2st_nearby   # From 0 introns
)) %>% unique()

# 3. Intergenic Information
# Since intergenic.info was originally built from input.data (which contains ALL candidates),
# we just need to format the IDs once.
Master_Intergenic <- intergenic.info %>%
  mutate(ids = paste0("intergenic_", ids)) %>%
  unique()

# 4. Create the final unified matching table (for the t.test later)
Master_final_stats_info <- do.call("rbind", list(
  Master_Intron_1st,
  Master_Intron_2nd,
  Master_Intergenic
)) %>% 
  dplyr::arrange(gene.id)

# 5. Create ONE deduplicated list of all introns for Cal_Coverage to save memory
Master_Unique_Introns <- do.call("rbind", list(Master_Intron_1st, Master_Intron_2nd)) %>%
  dplyr::select(chr, start, end, ids) %>% 
  dplyr::distinct(ids, .keep_all = TRUE)
```  

####====================#### 
Part5: Calculate all Intergenic and introns coverage for expressed tRNA-mRNA combination. Including genes with 1/0 intron
####====================####

### write intron and intergenic info into ``GRange`` object
```{r}
# intergenic
GRange_intergenic <- Master_Intergenic %>%
  dplyr::select(-strand) %>% 
  makeGRangesFromDataFrame(keep.extra.columns = TRUE,
                           ignore.strand = TRUE,
                           seqinfo = Seqinfo,
                           seqnames.field = "chr",
                           start.field = "start",
                           end.field = "end")
# Introns
GRange_Unique_Introns <- Master_Unique_Introns %>%
  makeGRangesFromDataFrame(keep.extra.columns = TRUE,
                           ignore.strand = TRUE,
                           seqinfo = Seqinfo,
                           seqnames.field = "chr",
                           start.field = "start",
                           end.field = "end")
```
## use ``Cal_Coverage()`` to look through ``bam`` files for coverage of each base
```{r, coverage of intergenic}
# initialize the function to calculate coverage
Cal_Coverage <- function(gr, bam){
  grl <- split(gr, f = gr$ids)
  cv <- lapply(grl, function(x){
    rng <- seq(start(x), end(x))
    chr <- seqnames(x)
    coverage(bam, param = ScanBamParam(which = x))[[chr]][rng]
  })
  cv
}

# calculate the AverageCoverage and nZero for each sample
coverage.intergenic <- lapply (bamfiles,function(bamfile){
  Cal_Coverage(GRange_intergenic, bamfile)
}) 

names(coverage.intergenic) <- SampleName
DicistronicName <- names(coverage.intergenic[[1]]) 
# give name of dicistronic in order, any sample is ok the order is the same 

# form a proper dataframe for all intergenic AC
AC.nZero.intergenic <- lapply(names(coverage.intergenic), function(x){
  coverage.intergenic[[x]] %>%
    lapply(function(i){
      data.frame(nZero = sum(i == 0), 
                 AveCov = mean(i))}) %>% 
    dplyr::bind_rows() %>% 
    set_colnames(c(paste0(x,"_nZero"),paste0(x,"_AveCov"))) 
}) %>%
  dplyr::bind_cols() %>%
  dplyr::mutate(ids = DicistronicName) %>%
  dplyr::select(ids,everything())

AC.intergenic <- AC.nZero.intergenic %>% 
  dplyr::select(ids, contains("AveCov"))

nZero.intergenic <- AC.nZero.intergenic %>% 
  dplyr::select(ids, contains("nZero"))
```

```{r, coverage of introns}
# calculate the AC and nZero for each sample
coverage.unique.introns <- lapply(bamfiles, function(bamfile){
  Cal_Coverage(GRange_Unique_Introns, bamfile) 
})
names(coverage.unique.introns) <- SampleName
UniqueIntronNames <- names(coverage.unique.introns[[1]])

# 3. Build the Dictionary (Lookup Table)
AC_Dictionary <- lapply(names(coverage.unique.introns), function(x){
  coverage.unique.introns[[x]] %>%
    lapply(function(i){ data.frame(AveCov = mean(i)) }) %>% 
    dplyr::bind_rows() %>%
    set_colnames(paste0(x,"_AveCov"))
}) %>%
  dplyr::bind_cols() %>%
  dplyr::mutate(intron_id = UniqueIntronNames) %>%
  dplyr::select(intron_id, everything())

# 4. Map the dictionary coverage back to your duplicated candidate list
AC.introns.mapped <- Master_final_stats_info %>%
  dplyr::filter(type %in% c("intron_closest_to_intergenic", "intron_2nd_closest_to_intergenic")) %>%
  dplyr::left_join(AC_Dictionary, by = c("ids" = "intron_id")) %>%
  dplyr::select(ids, contains("AveCov"))

# 5. create AC.final for your T-test
AC.final <- rbind(
  AC.intergenic,
  AC.introns.mapped
)
```

## match the intron to each tRNA-gene combo
```{r}
# Define the exact 13 columns in the exact order we want for everything
target_cols <- c("chr", "start", "end", "strand", "ids", "gene.id", "type", 
                 "gene.strand", "length", "id", "sameStrand", "tRNAFirst", "tRNA.upstream")
# 1. Create a "Dictionary" from your intergenic data
Matchmaker_Dict <- intergenic.info %>%
  dplyr::select(gene.id, id = ids, sameStrand, tRNAFirst, tRNA.upstream)

# 2. Attach the candidate IDs to ALL 1st Introns & enforce column order
Master_Intron_1st_mapped <- Master_Intron_1st %>%
  dplyr::select(-sameStrand, -tRNAFirst, -tRNA.upstream) %>% 
  dplyr::inner_join(Matchmaker_Dict, by = "gene.id") %>%
  dplyr::select(dplyr::all_of(target_cols))

# 3. Attach the candidate IDs to ALL 2nd Introns & enforce column order
Master_Intron_2nd_mapped <- Master_Intron_2nd %>%
  dplyr::select(-sameStrand, -tRNAFirst, -tRNA.upstream) %>%
  dplyr::inner_join(Matchmaker_Dict, by = "gene.id") %>%
  dplyr::select(dplyr::all_of(target_cols))

# 4. Format the Intergenic data & enforce column order
Master_Intergenic_mapped <- intergenic.info %>%
  dplyr::mutate(id = ids) %>%
  dplyr::mutate(ids = paste0("intergenic_", ids)) %>%
  dplyr::select(dplyr::all_of(target_cols))

# 5. Stack them together into the final Matchmaker table!
Master_final_stats_parid <- do.call("rbind", list(
  Master_Intron_1st_mapped,
  Master_Intron_2nd_mapped,
  Master_Intergenic_mapped
)) %>% 
  unique() %>%
  dplyr::arrange(id)
```

####====================#### 
Part5: Statistic test: Intergenic coverage > Intron coverage, introns coverage same between 1st and 2nd closest to intergenic region
####====================####

```{r}
final.stats.test <- Master_final_stats_parid %>%
  dplyr::select(chr, start, end, strand, ids, id, type, length)

aveCoverage.all <- AC.final %>%
  left_join(final.stats.test, by = "ids") %>%
  arrange(id) %>%
  unique()

aveCoverage.final <- aveCoverage.all %>%
  reshape2::melt(id.vars = c("ids", "chr", "start", "end", "type", "id", "strand", "length"),
       value.name = "aveCoverage", variable.name = "Sample") %>%
  mutate(Sample = gsub(x = .$Sample, pattern = "_AveCov", replacement = "")) %>%
  as_tibble() %>%
  arrange(id, Sample, type)
```
# statistical test
```{r}
# Map the tRNAs to genes
tRnaToGene <- aveCoverage.final %>%
  filter(grepl("t[rR][nN][aA]", ids),
         !is.na(aveCoverage)) %>%
  distinct(ids, .keep_all = TRUE) %>%
  dplyr::select(contains("id")) %>%
  mutate(id = as.character(id)) %>%
  split(f = .$id) %>%
  lapply(function(x){
    if (nrow(x) == 1) return(x) # This removes the genes with >1 tRNA, already fixed with unique tRNA-gene combo id
  }) %>%
  bind_rows() 
```

```{r}
# create a list with 2 elements with different comparison and do the t.test
tTestResults.final <- tRnaToGene$ids %>% 
  # extract(1) %>%
  lapply(function(x){
    # Get the geneid & coverage as a matrix
    combo <- filter(tRnaToGene, ids ==x)$id
    cov <- filter(aveCoverage.final, id == combo) %>%
      acast(Sample~type, value.var = "aveCoverage")
    
    #Initialise the output
    out <- data_frame(
      ids = character(),
      id = character(),
      comparison = character(),
      df = double(),
      V = double(),
      p = double()
    )
    
    # Compare the introns
    if (all(c("intron_2nd_closest_to_intergenic", "intron_closest_to_intergenic") %in% colnames(cov))) {
      tTest <- t.test(log(cov[,"intron_closest_to_intergenic"]+1),
                      log(cov[,"intron_2nd_closest_to_intergenic"]+1),
                      paired = TRUE)
      out %<>%
        bind_rows(
          data_frame(ids = x,
                     id = combo,
                     comparison = "intron_2nd_closest_to_intergenicVsintron_closest_to_intergenic",
                     df = tTest$parameter, # degree of freedom
                     V = tTest$statistic,
                     p = tTest$p.value)
        )
    }
    
    # Compare the intergenic regions
    if (all(c("intergenic", "intron_closest_to_intergenic") %in% colnames(cov))) {
      tTest <- t.test(log(cov[,"intron_closest_to_intergenic"]+1),
                      log(cov[,"intergenic"]+1),
                      paired = TRUE)
      out %<>%
        bind_rows(
          data_frame(ids = x,
                     id = combo,
                     comparison = "IntergenicVsintron_closest_to_intergenic",
                     df = tTest$parameter,
                     V = tTest$statistic,
                     p = tTest$p.value)
        )
    }
    out
    
  }) %>%
  bind_rows() %>%
  mutate(adjP = p.adjust(p, "bonferroni"),
         FDR = p.adjust(p, "fdr")) %>%
  arrange(p) %>%
  split(f = .$comparison)
```
## t.test output
```{r two way filtering}
# for FDR
intergeic.sign.final <- tTestResults.final$IntergenicVsintron_closest_to_intergenic %>% 
  filter(FDR < params$fdr_threshold, V < 0)
intron.not.sign.final <- tTestResults.final$intron_2nd_closest_to_intergenicVsintron_closest_to_intergenic %>% 
  filter((FDR < params$fdr_threshold & V > 0) | FDR >= params$fdr_threshold)
final.sign <- intersect(intron.not.sign.final$id,intergeic.sign.final$id)
```

####====================#### 
Part6: Assess continuous coverage for candidates that pass the t.test
####====================####

```{r}
allIntergenic <- names(coverage.intergenic[[1]]) 
allSamples <- names(coverage.intergenic)

coverage.samples <- lapply(allIntergenic, function(intergenic){
  # cbind is faster and cleaner than rbind + t() for building this matrix
  v <- lapply(allSamples, function(sample){
    as.integer(coverage.intergenic[[sample]][[intergenic]])
  }) 
  v <- do.call(cbind, v)
  colnames(v) <- allSamples
  return(v)
})
names(coverage.samples) <- allIntergenic

# 2. Filter our new Master list by the candidates that passed the T-test
coverage.list.final <- nZero.intergenic %>% 
  # Use inner_join on the unified master table
  dplyr::inner_join(Master_final_stats_parid, by = "ids") %>% 
  # Filter by the significant dicistronic candidate IDs
  dplyr::filter(id %in% final.sign) %>%
  # Make sure we are only looking at the intergenic rows
  dplyr::filter(type == "intergenic") %>%
  dplyr::mutate(continu.cov = rowSums(dplyr::select(., contains("nZero"))) == 0)

# 3. Assess the combined continuous coverage for the significant candidates
keptIntergenic <- coverage.list.final$ids

combined.nZero <- coverage.samples[names(coverage.samples) %in% keptIntergenic] %>% 
  lapply(function(mat){
    # Count bases where the sum of coverage across all samples is exactly 0
    sum(rowSums(mat) == 0)
  }) %>% 
  unlist() %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column("ids") %>%
  set_colnames(c("ids", "Combined.nZero"))

# 4. Attach the gap counts and create the final logical check
continuous.coverage.final <- coverage.list.final %>%
  dplyr::left_join(combined.nZero, by = "ids") %>%
  dplyr::mutate(combined.continu.cov = (Combined.nZero == 0))
```
####====================####
# Part 9: Final DT candidate list with tRNA, Gene, and CDS coordinates
####====================####
```{r}
# 1. Calculate the exact number of introns for every gene dynamically
Intron_Counts <- intron.info %>%
  dplyr::group_by(gene.id) %>%
  dplyr::summarise(num.Intron = dplyr::n(), .groups = "drop")

# 2. Add Intron count to your unified continuous coverage table
continuous.coverage.final_Nintrons <- continuous.coverage.final %>%
  dplyr::left_join(Intron_Counts, by = "gene.id") %>%
  dplyr::mutate(num.Intron = tidyr::replace_na(num.Intron, 0)) %>%
  dplyr::filter(combined.continu.cov == TRUE)

# 3. Combine with tRNA coordinates
Dicistronic.all.tRNA <- continuous.coverage.final_Nintrons %>%
  dplyr::left_join(tRNA.CDS.expressed.combo, by = "id") %>%
  dplyr::select(id, chr = chr.x, start, end, gene.id = gene.id.x, 
                tRNA.id, tRNA.start, tRNA.end, tRNA.strand, 
                tRNA.upstream = tRNA.upstream.x, tRNAFirst = tRNAFirst.x, 
                interval.length, num.Intron, combined.continu.cov,
                dplyr::contains("_nZero"))

# 4. Add Gene AND CDS boundary information
DT_with_Gene_Startend <- Dicistronic.all.tRNA %>%
  # Join overall Gene boundaries
  dplyr::left_join(gene.information, by = "gene.id") %>%
  dplyr::rename(
    DT.start    = start.x,
    DT.end      = end.x,
    gene.start  = start.y,
    gene.end    = end.y,
    gene.chr    = Chr,
    gene.strand = strand
  ) %>%
  dplyr::select(-type) %>%
  # Join CDS boundaries (Required for your classification matrix below)
  dplyr::left_join(CDS.start.end.for.gene %>% dplyr::select(gene.id, CDS.start = gene.start, CDS.end = gene.end), 
                   by = "gene.id")

# 5. estimate tRNA in UTR or intergenic
DT_Final_Classified <- DT_with_Gene_Startend %>%
  dplyr::mutate(
    tRNA_Overlaps_Gene = (tRNA.start <= gene.end & tRNA.end >= gene.start),
    tRNA_In_CDS       = (tRNA.start >= CDS.start & tRNA.end <= CDS.end),
    tRNA_Overlaps_CDS = (tRNA.start <= CDS.end & tRNA.end >= CDS.start),
    Location_Type = dplyr::case_when(
      tRNA_In_CDS == TRUE ~ "CDS-Internal",
      tRNA_Overlaps_CDS == TRUE & tRNA_In_CDS == FALSE ~ "CDS-Edge_Conflict",
      tRNA_Overlaps_Gene == TRUE & tRNA_Overlaps_CDS == FALSE ~ "UTR",
      TRUE ~ "Intergenic"
    )
  )
```
# export results
```{r}
write_xlsx(DT_Final_Classified, path = file.path(params$out_dir, "Final_Result_Manual_optimized.xlsx"))
```

####======= Finish DT detection for any candidates with any number of introns ========####